# Results


```{r}

#Load in all 22 datasets for single females and males from 2009-2019
library(tidyverse)
# install.packages("janitor")
library(janitor)


files <- list.files("/Users/snehajaikumar/EDAV/final_datasets", pattern = "*.xlsx", full.names = TRUE)
manual_header <- c(
  "Item",
  "All",
  "Under 25 years",
  "25-34 years",
  "35-44 years",
  "45-54 years",
  "55-64 years",
  "65 years and older"
)

clean_numeric <- function(x) {
  x |>
    str_replace("^[abc]/\\s*", "") |>   # remove a/, b/, c/ and any space after
    na_if("") |>                       # empty string → NA
    as.numeric()
}


read_cex_file <- function(filepath) {

  # Extract gender + year using regex
  gender <- if (str_detect(filepath, "females")) "Female" else "Male"
  year   <- str_extract(filepath, "\\d{4}") |> as.numeric()

  # Read raw
  raw <- read_excel(filepath, col_names = FALSE)

  # Extract row 3 as header
  # header <- as.character(raw[2, ])
  # header[header == "" | is.na(header)] <- paste0("unknown_", seq_len(sum(header == "" | is.na(header))))
  names(raw) <- manual_header

  # Add metadata
   df <- raw |>
      slice(47:229) |>       # <— keep only rows 47–183
      janitor::remove_empty("cols") |>
      filter(!(Item != "" & if_all(-Item, is.na))) |> filter(Item != "Average annual expenditures")
      # filter(!if_all(everything(), is.na)) |>   # <— remove fully-NA rows
   
   gift_start <- which(df$Item %in% c("Gifts of goods and services",
                                   "Gifts of goods and services, total"))
   gift_end   <- which(df$Item == "All other gifts")
   
   

  if (length(gift_start) == 1 && length(gift_end) == 1) {
    df <- df[-(gift_start:gift_end), ]
  }
   
    df <- df |> mutate(
        gender = gender,
        year = year
      ) |>
     mutate(across(-c(Item, gender, year), clean_numeric)) 
  

  return(df)
}

df_main <- map_df(files, read_cex_file)

# gift_start <- which(df_all$Item == "Gifts of goods and services")
# gift_end   <- which(df_all$Item == "All other gifts")
# df_gifts <- df_all[gift_start:gift_end, ]
# df_main <- df_all[-(gift_start:gift_end), ]

df_main



```


```{r}
library(dplyr)
library(tidyr)

df_long <- df_main |>
  pivot_longer(
    cols = -c(Item, gender, year),  
    names_to = "age_group",
    values_to = "expenditure"
  ) |>
  mutate(
    age_group = factor(
      age_group,
      levels = c(
        "All",
        "Under 25 years",
        "25-34 years",
        "35-44 years",
        "45-54 years",
        "55-64 years",
        "65 years and older"
      ),
      ordered = TRUE
    )
  )

df_missing <- df_long

df_long <- df_long |>
  mutate(
    Item = case_when(
      Item %in% c("Gasoline, other fuels, and motor oil",
                  "Gasoline and motor oil") ~ "Gasoline & motor oil",
      Item %in% c("Public transportation",
                  "Public and other transportation") ~ "Public transportation",
      Item %in% c("Males, 2 and over",
                  "Men and boys") ~ "Men and boys",
      Item %in% c("Boys, 2 to 15", "Men, 16 and over") ~ "Men and boys",
      Item %in% c("Females, 2 and over",
                  "Women and girls") ~ "Women and girls",
      Item %in% c("Girls, 2 to 15", "Women, 16 and over") ~ "Women and girls",
      
      TRUE ~ Item
    )
  )

df_long

```
```{r}
df_hi <- df_long |>
  filter(Item %in% c("Housing", "Income after taxes"))

df_hi

df_hi_wide <- df_hi |>
  pivot_wider(
    names_from = Item,
    values_from = expenditure
  )

df_hi_wide


df_hi_wide <- df_hi_wide %>%
  mutate(housing_burden = Housing / `Income after taxes`) #|> mutate(year = factor(year))

df_hi_wide

```
```{r}
ggplot(df_hi_wide, aes(x = year, y = housing_burden, color = gender)) +
  geom_line(size = 1.2) +
  facet_wrap(~ age_group) +
  scale_y_continuous(labels = scales::percent) +
  scale_x_continuous(breaks = seq(min(df_hi_wide$year), max(df_hi_wide$year), 1)) +
  labs(
    title = "Housing Burden Over Time (2009–2019)",
    y = "% of Income Spent on Housing",
    x = "Year"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )
```

```{r}
major_groups_main <- c(
  "Food", "Housing", "Apparel and services", "Transportation", "Healthcare",
   "Education","Cash contributions",
  "Personal insurance and pensions","Personal care products and services"
)

df_major <- df_long |> 
  filter(Item %in% major_groups_main)

df_gender_wide <- df_major |>
  pivot_wider(
    names_from = gender,
    values_from = expenditure
  )

df_gender_wide <- df_gender_wide |>
  mutate(
    gender_diff = Female - Male    # positive = women pay more
  )

ggplot(df_gender_wide |> filter(age_group == "All"),
       aes(x = year, y = gender_diff, color = Item)) +
  geom_line(size = 1, color = "steelblue") +
  facet_wrap(~ Item, scales = "free_y") +
  labs(
    title = "Gender Difference in Annual Spending (Female − Male)",
    subtitle = "2009–2019, Single Adults",
    x = "Year",
    y = "Difference in Spending ($)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  scale_x_continuous(breaks = seq(2009, 2019, 1))


```

```{r}

df_major_all <- df_major |> filter(age_group == "All")

df_mean <- df_major_all |>
  group_by(gender, Item) |>
  summarise(mean_expenditure = mean(expenditure, na.rm = TRUE), .groups = "drop")

df_diff <- df_mean |>
  pivot_wider(
    names_from = gender,
    values_from = mean_expenditure
  )

df_diff <- df_diff |>
  mutate(diff = Female - Male)

df_diff <- df_diff |> rename(level1 = Item)


ggplot(df_diff,
       aes(x = reorder(level1, diff),
           y = diff,
           fill = diff > 0)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "firebrick", "FALSE" = "steelblue")) +
  labs(
    title = "Gender Spending Difference (Female − Male), Average 2009–2019",
    subtitle = "Positive values indicate categories where women spend more",
    x = "Category",
    y = "Difference in Average Annual Spending ($)",
    fill = ""
  ) +
  theme_minimal() +
  theme(legend.position = "none")


```
```{r}
transport_items <- c(
  "Cars and trucks, new",
  "Cars and trucks, used",
  "Other vehicles",
  "Gasoline & motor oil",
  "Vehicle finance charges",
  "Maintenance and repairs",
  "Vehicle rental, leases, licenses, and other charges",
  "Vehicle insurance",
  "Public transportation"
  
)

big_palette <- c(
  "#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e",
  "#e6ab02", "#a6761d",  "#6a3d9a","#1f78b4"
)


df_transport <- df_long |>
  filter(Item %in% transport_items) |>
  filter(age_group == "All")

df_transport_summary <- df_transport |>
  group_by(year, gender, Item) |> 
  summarise(exp = mean(expenditure, na.rm = TRUE),
            .groups = "drop")

ggplot(df_transport_summary,
       aes(x = year, y = exp, color = Item)) +
  
  geom_line(size = 1.0, alpha = 0.9) +        # ← LINES
  geom_point(size = 2.2) +                    # ← DOTS ON EVERY POINT
  
  scale_color_manual(values = big_palette) +  # your 20-color palette
  
  facet_wrap(~ gender) +
  
  labs(
    title = "Transportation Spending Breakdown by Gender (2009–2019)",
    subtitle = "Lines show subcategory trends; dots show observed values",
    y = "Annual Spending ($)",
    x = "Year",
    color = "Category"
  ) +
  
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.key.size = unit(0.5, "cm"),
    legend.position = "right"
  ) +
  
  scale_x_continuous(breaks = seq(2009, 2019, 1))



target <- "Public transportation"

ggplot(df_transport_summary, aes(x = year, y = exp, group = Item)) +

  # 1. BASE LINES — dimmed for all non-target categories
  geom_line(
    data = df_transport_summary |> filter(Item != target),
    aes(color = "other"),
    size = 0.7,
    alpha = 0.35
  ) +

  # 2. TARGET LINE — bold & colored
  geom_line(
    data = df_transport_summary |> filter(Item == target),
    aes(color = target),
    size = 1.4
  ) +

  # 3. TARGET POINTS — emphasize observations
  geom_point(
    data = df_transport_summary |> filter(Item == target),
    aes(color = target),
    size = 3
  ) +

  facet_wrap(~ gender) +
  
  scale_x_continuous(breaks = seq(2009, 2019, 1)) +

  # 4. MANUAL COLORS
  scale_color_manual(
    name = "Category",
    values = c(
      "other" = "gray70",
      target = "#E41A1C"   # bold red highlight
    ),
    labels = c(
      "other" = "Other transportation categories",
      target = target
    )
  ) +

  labs(
    title = "Transportation Spending Breakdown (Highlighting Public Transportation)",
    subtitle = "Public transportation emphasized for clearer gender trends",
    x = "Year",
    y = "Annual Spending ($)"
  ) +

  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )



```
```{r}
df_pc_income <- df_long |>
  filter(
    Item %in% c("Personal care products and services", "Income after taxes"),
  )


df_pc_income_wide <- df_pc_income |>
  select(year, gender, age_group, Item, expenditure) |>
  pivot_wider(
    names_from = Item,
    values_from = expenditure
  )

df_pc_income_wide <- df_pc_income |>
  select(year, gender, age_group, Item, expenditure) |>
  pivot_wider(
    names_from = Item,
    values_from = expenditure
  )

df_pc_income_wide <- df_pc_income_wide |>
  mutate(
    pct_budget_pc = `Personal care products and services` / `Income after taxes`
  )


ggplot(df_pc_income_wide,
       aes(x = year, y = pct_budget_pc, color = gender)) +

  geom_line(size = 1.2) +
  geom_point(size = 2.8) +

  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  scale_x_continuous(breaks = seq(2009, 2019, 1)) +

  scale_color_manual(values = c("Female" = "#E41A1C", "Male" = "#377EB8")) +

  facet_wrap(~ age_group) +

  labs(
    title = "Personal Care Spending as a % of Income After Taxes",
    subtitle = "By Gender and Age Group (2009–2019)",
    x = "Year",
    y = "Percent of After-Tax Income",
    color = "Gender"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(size = 11, face = "bold")
  )


```
```{r}

apparel_items <- c(
  "Men and boys",
  "Women and girls",
  "Children under 2"
)


df_apparel_small <- df_long |>
  filter(
    Item %in% apparel_items,
    age_group %in% c("Under 25 years", "25-34 years")
  )

df_apparel_small$age_group <- factor(
  df_apparel_small$age_group,
  levels = c("Under 25 years", "25-34 years"),
  ordered = TRUE
)

df_apparel_small_combined <- df_apparel_small |>
  group_by(Item, gender, year, age_group) |>
  summarise(
    expenditure = sum(expenditure, na.rm = TRUE),
    .groups = "drop"
  )


df_apparel_small_combined


df_female <- df_apparel_small_combined |> filter(gender == "Female")

ggplot(df_female,
       aes(x = year, y = expenditure, color = Item)) +
  geom_line(size = 1.1) +
  geom_point(size = 2) +
  facet_wrap(~ age_group, scales = "free_y") +
  labs(
    title = "Female Apparel Spending Over Time",
    subtitle = "Under 25 and 25–34 age groups (2009–2019)",
    x = "Year",
    y = "Annual Spending ($)",
    color = "Apparel Category"
  ) +
  scale_color_brewer(palette = "Dark2") +
  scale_x_continuous(breaks = seq(2009, 2019, 1)) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

df_male <- df_apparel_small_combined |> filter(gender == "Male")

ggplot(df_male,
       aes(x = year, y = expenditure, color = Item)) +
  geom_line(size = 1.1) +
  geom_point(size = 2) +
  facet_wrap(~ age_group, scales = "free_y") +
  labs(
    title = "Male Apparel Spending Over Time",
    subtitle = "Under 25 and 25–34 age groups (2009–2019)",
    x = "Year",
    y = "Annual Spending ($)",
    color = "Apparel Category"
  ) +
  scale_color_brewer(palette = "Dark2") +
  scale_x_continuous(breaks = seq(2009, 2019, 1)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))



ggplot(df_apparel_small_combined,
       aes(x = year, y = expenditure,
           color = gender, linetype = gender)) +
  geom_line(size = 1.1) +
  geom_point(size = 2) +
  facet_wrap(~ Item + age_group, scales = "free_y") +
  labs(
    title = "Male vs Female Apparel Spending, Side-by-Side",
    subtitle = "Each panel = one apparel subcategory × one age group",
    x = "Year",
    y = "Annual Spending ($)",
    color = "Gender",
    linetype = "Gender"
  ) +
  scale_color_manual(values = c(Female = "#E41A1C", Male = "#377EB8")) +
  scale_x_continuous(breaks = seq(2009, 2019, 1)) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, margin = margin(t = 10)),
    strip.text = element_text(face = "bold")
  )


```




```{r}
library(dplyr)
library(vcd)

health_items <- c(
  "Health insurance",
  "Medical services",
  "Drugs",
  "Medical supplies"
)

gender_colors <- c(Female = "#E41A1C", Male = "#377EB8")

#FIRST YEAR
df_mosaic_1 <- df_long |>
  filter(
    year == 2009,
    age_group == "All",
    Item %in% health_items
  ) |>
  select(Item, gender, expenditure) |>
  rename(Freq = expenditure)

# 1) Turn into a table of dollar amounts
tab <- xtabs(Freq ~ Item + gender, data = df_mosaic_1)

rownames(tab) <- recode(
  rownames(tab),
  "Health insurance" = "Insurance",
  "Medical services" = "   Serv",
  "Medical supplies" = "Supp"
)

mosaic(
  tab,
  labeling = labeling_values,
  labeling_args = list(
    labels = tab,   
    gp_text = gpar(cex = 0.9, col = "black")
  ),
  highlighting = "gender",
  highlighting_fill = gender_colors,
  direction = c("h", "v"),
  main = "Healthcare Spending by Gender (2009)",
  sub  = "Supp = Medical Supplies, Serv = Medical Services"
)



#SECOND YEAR
df_mosaic_2 <- df_long |>
  filter(
    year == 2016,
    age_group == "All",
    Item %in% health_items
  ) |>
  select(Item, gender, expenditure) |>
  rename(Freq = expenditure)

# 1) Turn into a table of dollar amounts
tab <- xtabs(Freq ~ Item + gender, data = df_mosaic_2)

rownames(tab) <- recode(
  rownames(tab),
  "Health insurance" = "Insurance",
  "Medical services" = "   Serv",
  "Medical supplies" = "Supp"
)

mosaic(
  tab,
  labeling = labeling_values,
  labeling_args = list(
    labels = tab,      
    gp_text = gpar(cex = 0.9, col = "black")
  ),
  highlighting = "gender",
  highlighting_fill = gender_colors,
  direction = c("h", "v"),
  main = "Healthcare Spending by Gender (2016)",
  sub  = "Supp = Medical Supplies, Serv = Medical Services"
)

#THIRD YEAR
df_mosaic_3 <- df_long |>
  filter(
    year == 2019,
    age_group == "All",
    Item %in% health_items
  ) |>
  select(Item, gender, expenditure) |>
  rename(Freq = expenditure)

# 1) Turn into a table of dollar amounts
tab <- xtabs(Freq ~ Item + gender, data = df_mosaic_3)

rownames(tab) <- recode(
  rownames(tab),
  "Health insurance" = "Insurance",
  "Medical services" = "   Serv",
  "Medical supplies" = "Supp"
)

mosaic(
  tab,
  labeling = labeling_values,
  labeling_args = list(
    labels = tab,        
    gp_text = gpar(cex = 0.9, col = "black")
  ),
  highlighting = "gender",
  highlighting_fill = gender_colors,
  direction = c("h", "v"),
  main = "Healthcare Spending by Gender (2019)",
  sub  = "Supp = Medical Supplies, Serv = Medical Services"
)


```

```{r}
df_missing
df_missing |>
  filter(is.na(expenditure)) |>
  count(Item, sort = TRUE) |>
  slice_head(n = 20) |>
  ggplot(aes(x = reorder(Item, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Items With the Most Missing Expenditure Values",
    x = "Item",
    y = "Number of Missing Values"
  ) +
  theme_minimal()



library(tidyverse)

top_items <- df_missing |>
  filter(is.na(expenditure)) |>
  count(Item, sort = TRUE) |>
  slice_head(n = 20) |>
  pull(Item)

df_missing |>
  filter(Item %in% top_items) |>
  mutate(is_missing = ifelse(is.na(expenditure), 1, 0)) |>
  count(Item, gender, wt = is_missing) |>
  ggplot(aes(x = gender, y = reorder(Item, n), fill = n)) +
  geom_tile(color = "white") +
  scale_fill_gradientn(
    colours = c("#DEEBF7","#9ABBDD","#4A7BBF","#08306B"),
    values = scales::rescale(c(0, 1, 5, max(df_heat$n_missing))),
    name = "Missing Count"
  )  +
  labs(
    title = "Heatmap of Missing Values by Item and Gender (Top 20 Items)",
    x = "Gender",
    y = "Item",
    fill = "Missing\nCount"
  ) +
  theme_minimal(base_size = 13)



library(tidyverse)

top_items <- df_missing |>
  filter(is.na(expenditure)) |>
  count(Item, sort = TRUE) |>
  slice_head(n = 20) |>
  pull(Item)

df_heat_item_age <- df_missing |>
  filter(Item %in% top_items) |>
  mutate(is_missing = ifelse(is.na(expenditure), 1, 0)) |>
  count(Item, age_group, wt = is_missing) |>
  mutate(n = as.integer(n)) |>
  complete(Item, age_group, fill = list(n = 0))

ggplot(df_heat_item_age,
       aes(x = age_group, y = reorder(Item, n), fill = n)) +
  geom_tile(color = "white") +
  scale_fill_gradient(
    low = "#deebf7",   # light blue
    high = "#08306b",
    name = "Missing\nCount"
  ) +
  labs(
    title = "Missing Expenditure Values by Item and Age Group",
    x = "Age Group",
    y = "Item"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))





```


```{r}

df_heat <- df_long |>
  mutate(is_missing = ifelse(is.na(expenditure), 1, 0)) |>
  group_by(age_group, gender) |>
  summarise(n_missing = sum(is_missing), .groups = "drop") |>
  complete(age_group, gender, fill = list(n_missing = 0))



ggplot(df_heat, aes(x = age_group, y = gender, fill = n_missing)) +
  geom_tile(color = "white") +
  scale_fill_gradient(
    low = "#deebf7",   # light blue
    high = "#08306b",
    name = "Missing Count"
  )  +
  labs(
    title = "Missing Values by Age Group and Gender",
    x = "Age Group",
    y = "Gender",
    #fill = "Missing Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

We noticed that healthcare spending always had positive values for the spending gap and since healthcare spending is inevitable, it makes it an uncontrollable burden, which is why we wanted to delve deeper into the healthcare cost burden by gender. 

Firstly, to visualize if there was a cost burden on the healthcare spending, we used CMS.gov’s Total per-capita Health Expenditure by Age and Sex dataset (including Medicare, Medicaid, Private Health Insurance, and Out of pocket) to plot horizontal bar charts faceted over years using the difference in values (Female - Males). We found that most of the bars were positive for all the age groups except for 65-84, and 0-18. This means that most women have been spending more than men for years. 

